Algebraic Tori in Unitary Groups

### 2014-04-29 Paper.tex
Added a bit more

@@

@@

For our purposes a unitary group is a simply connect group $G$ of type $A_n$.
We may construct such a group as follows.
Fix the following:
\begin{itemize}
\item An quadratic \'etale algebra $E$ over $K$.
\item A central simple $E$-algebra $A$ of degree $n$ with anti-involution $\sigma$ restricting to the non-trivial involution of $E$.
\item An element $J\in F^\sigma$ of norm $\delta$ whose minimal polynomial at a real place $\nu$ of $K$ which ramifies in $E$ has $r_\nu$ positive roots and $s_\nu$ negative roots.
\end{itemize}
Define then $G$ to be the group scheme whose points over a $K$-algebra $R$ are given by:
\[ G(R) = \{ g\in (A\otimes_K R)^\times \mid \sigma(g)Jg = J \text{ and } Nm(g) = 1 \}. \]

We claim that all $G$ arise this way, and that if $n>1$ then $G$ is uniquely determined by the invariants $(E,A,\delta, \left\vert r_\nu-s_\nu\right\vert)$ where $\delta \in K^\times/(K^\times)^nN_{E/K}(E^\times)$ subject only to the condition that $(-1)^{r_\nu-s_\nu}\delta$ is positive at all the real places of $K$ which ramify in $E$.

@@

\begin{remark}
It should be pointed out that the group $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is trivial if either $n$ is odd or if $E = K\times K$ is split.
Further, for local fields at least one of $K^\times/(K^\times)^nN_{E/K}(E^\times)$ or $H^2(\GalKbK, T_{E,\sigma}) $ is trivial.
\end{remark}

@@

\begin{remark}
There are several subtly different ways to define the Hasse (or Witt or Hasse-Witt) invariant, these typically differ by multiplication in the Brauer group by a constant depending on the discriminant.
The one we have given is not equal to the coboundary map for the exact sequence above.
\end{remark}

@@

We claim that all orthogonal groups arise this way (note that $\hat{G}$ admits further forms in the case $n=4$).

@@

By \cite[29.16]{TODO - The book of involutions}
the cohomology set $H^1(\Gamma, Isom(A,\tau)) = H^1(\Gamma,\tilde{G}) $ classifies the '$\tau$-symmetric elements of $A^\times$' up to change of basis.
The action of $H^1(\GalKbK,\mu_2) $ is to rescale these by an element of $k^\times/(k^\times)^2$.
\begin{remark}
When $A$ is split, this is nothing other than the classification of quadratic forms up to rescaling as above.
In even dimensions, this depends on the discriminant and the base change of the Hasse invariant to `discriminant' algebra.

When $A$ is a quaternion algebra, the analysis proceeds as above, and we may still associate to $G$ a discriminant, a quaternion algebra and signature information. 
As we shall not need it in the sequel, we omit the details, but remark that among other simplifications, $J$ can be taken to be diagonal, and the diagonal entries can be rescaled by $(K^\times)^2$
\end{remark}

@@

>>>>

\TODO{Many well known results are still stated without proof here}

@@

\subsection{Key Results in Galois Cohomology}

We shall make frequent use of the following results of Galois cohomology.

\TODO{Hasse Principle} The map from Global to local is injective.
\TODO{Simply Connected Local Fields Thing} $H^1(G)$ is trivial if $G$ simply connected at a non-real local field
\TODO{Simply connected real fields results} I feel someone has given an interpretation in terms of "indexes" of groups...

optional:
\TODO{H1 classifies torsors}
\TODO{remarks re twisting}
\TODO{H1 and H2 for relevant tori, restrictions of scalars and norms} (possibly in a later section?)

@@

For our purposes a unitary group is a simply connect group $G$ of type $A_n$.
We may construct such a group as follows.
Fix the following:
\begin{itemize}
\item An quadratic \'etale algebra $E$ over $K$.
\item A central simple $E$-algebra $A$ of degree $n$ with anti-involution $\sigma$ restricting to the non-trivial involution of $E$.

         Though unnecissary at this point, we note that for simplicity one is free to suppose that $\sigma$ has discriminant $1$ and is positive at all the real places of $K$ which ramify in $E$.
         By the disciminant of $\sigma$ we mean the norm from $A$ to $K$ of an element $a$ with $\sigma(a)=-a$, by \cite[TODO]{book of involutions} this is well defined up to $N_{E/K}(E^\times)(K^\times)^n$. To say the involution is positive we mean that $a\sigma(a) \ge 0$.

\item An element $J\in A^\sigma$ of norm $\delta$ whose minimal polynomial at a real place $\nu$ of $K$ which ramifies in $E$ has $r_\nu$ positive roots and $s_\nu$ negative roots.
\end{itemize}
Define then $G$ to be the group scheme whose points over a $K$-algebra $R$ are given by:
\[ G(R) = \{ g\in (A\otimes_K R)^\times \mid \sigma(g)Jg = J \text{ and } Nm(g) = 1 \}. \]

\begin{theorem}
All forms of $G$ arise this way. Moreover, if $n>1$ then $G$ is uniquely determined by the invariants $(E,A,\delta, \left\vert r_\nu-s_\nu\right\vert)$ where $\delta \in K^\times/(K^\times)^nN_{E/K}(E^\times)$ subject only to the condition that $(-1)^{r_\nu-s_\nu}\delta$ is positive at all the real places of $K$ which ramify in $E$.
\end{theorem}
\cite[TODO]{book of involutions}

The cohomological interpretation of the above theorem is as follows:

@@

\begin{remark}
It should be pointed out that the group $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is trivial if either $n$ is odd or if $E = K\times K$ is split.
Further, for local fields at least one of $K^\times/(K^\times)^nN_{E/K}(E^\times)$ or $H^2(\GalKbK, T_{E,\sigma}) $ is trivial. In the case of $H^2(\GalKbK, T_{E,\sigma}) $ trivial, the algebra $A$ is the matrix algebra, and the element of $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is giving the discriminant of the involution, equivalently, the discriminant of the Hermitian space.
\end{remark}

@@

\begin{remark}
There are several subtly different ways to define the Hasse (or Witt or Clifford) invariant, these typically differ by multiplication in the Brauer group by a constant depending on the discriminant.
The one we have given (the standard definition of the Hasse invariant) is not equal to the coboundary map for the exact sequence above.
\end{remark}

@@

\begin{theorem}
All forms of $G$ arise this way. (note that $\hat{G}$ admits further forms in the case $n=4$).
\end{theorem}
\cite[TODO]{book of involutions}

The cohomological interpretation of this result is as follows:

@@

By \cite[29.16]{TODO - The book of involutions}
the cohomology set $H^1(\Gamma, Isom(A,\tau)) = H^1(\Gamma,\tilde{G}) $ classifies the '$\tau$-symmetric elements of $A^\times$' up to change of basis.
The action of $H^1(\GalKbK,\mu_2) $ is to rescale these by an element of $k^\times/(k^\times)^2$.
\begin{remark}
When $A$ is split, this is nothing other than the classification of quadratic forms up to rescaling as above.
In even dimensions, this depends on the discriminant and the base change of the Hasse invariant to the `discriminant' algebra.

When $A$ is a quaternion algebra, the analysis proceeds as above, and we may still associate to $G$ a discriminant, a central simple algebra (which is $2$-torsion if $n$ is even, and $4$-torsion when $n$ is odd) and signature information. 
As we shall not need it in the sequel, we omit the details, but remark that among other simplifications, $J$ can be taken to be diagonal, and the diagonal entries can be rescaled by $(K^\times)^2$
\end{remark}

@@

\subsection{Classification of Tori in Unitary Groups}

\begin{theorem}
Let $(E,A,\tau)$ be a triple consisting of a quadratic \'etale algebra $E$, a degree $n$ central simple algebra $A$ over $E$ with an involution $\tau$ of $A$ which restricts to the unique non-trivial involution on $E$. Let $U_{E,A,\tau}$ be the associated unitary group. Let $\delta_{\tau} \in K^\times/N_{E/K}(E^\times)$ denote the discriminant of $\tau$.

Let $(E,M_n,\tilde{\tau})$ be the triple consisting of $E$, the matrix algebra of degree $n$ over $E$, and an involution $\tilde{\tau}$ of disciminant $\delta_{\tau}$ so that $U_{E,A,\tau}$ is the unitary group for the $E$-Hermitian space of discriminant $\delta_{\tau}$.

The rational conjugacty classes of maximal tori in $U_{E,A,\tau}$ are naturally identified with a subset of those of $U_{E,M_n,\tilde{\tau}}$.
The tori in $U_{E,M_n,\tilde{\tau}}$ are assicated to triples $(F,\sigma,\lambda)$, the rational conjugacy classes of tori in $U_{E,A,\tau}$ are precisely the subset for which $(F,\sigma)\hookrightarrow (A,\tau)$, equivalently, the subset for which $F$ splits $A$.

The tori associated to $(F,\sigma,\lambda)$ in $U_{E,A,\tau}$ is the same as that in $U_{E,M_n,\tilde{\tau}}$ in particular they are $T_{F,\sigma}$, fixing an embedding $\rho : F \rightarrow A$, and an involution $\hat\tau$ of disciminant $1$ the conjugacy classes correspond to the image of $T_{F,\sigma}$ in $U_{E,A,\tau} \subset A$ where $U_{E,A,\tau}$ is defined comes from $\tau(x) = \rho(\lambda^{-1})\hat\tau(x)\rho(\lambda)$.

The tori in $SU_{E,A,\tau}$ are the subtori $T_{F,\sigma,N_{F/E}}$.
\end{theorem}
\begin{proof}
The proof is clear locally in terms of existence. 
(Cite a result re embedding of algebras).

Local existance -> global existance is easy, as $\lambda \in F$ places $\lambda \in A$ so as to allow us to define the involution, which has the correct local invariants and thus gives the correct unitary group.

Local conjugacy class statement requires knowing conjugacy classes in the local non-split algebra setting. 
The exact sequence $H^1(F) \rightarrow H^1(Aut(A,F)) \rightarrow H^1(Aut(F))$ tells us these are all conjugate.
A direct argument of this claim would be nice.
(Conjugacy of maximal \'etale sub-algebras of central simple algebras must be known).

Global conjugacy class statement is then only claiming there are not more conjugacy classes than this, as global conjugacy would imply local conjugacy.
It is "clear" they are conjugate if they differ by a norm of $F$, because this element can be used to "change" basis.
(This is basically the Hasse principal in this case).
\end{proof}

\begin{remark}
The cohomological interpretation of this result is as follows.

We have shows already that rational conjugacy classes of tori in $U_{E,A,\tau}$ correspond to distinguished subalgebras with involution $(F,\sigma)$.
Following the book of involutions (TODO-cite), we may classify pairs of $((E,A,\tau),(F,\sigma))$ by $H^1(Aut_K((E,A,\tau),(F,\sigma)))$.

There is an exact sequence:
\[ 1 \rightarrow T_{F,\sigma} \rightarrow Aut_K((E,A,\tau),(F,\sigma)) \rightarrow Aut_K(F,\sigma) \rightarrow 1 \]
An element of $H^1(Aut_K(F,\sigma))$ specifies an \'etale algebra $F$, the element of $H^1(T_{F,\sigma})$ determines the conjugacy class.

The elements of $H^1( Aut_K((E,A,\tau),(F,\sigma)))$ giving tori in $U_{E,A,\tau}$ is precisely the kernel of the map to $H^1(Aut_K((E,A,\tau))$.

If we choose to work as in  (TODO cite reeder) with pure inner forms,
We set $N$ to be the normalizer in $U_{E,A,\tau}$ of $T_{F,\sigma}$ and $Z$ be the center of $U_{E,A,\tau}$, we obtain an exact sequence:
\[ 1 \rightarrow Z \rightarrow N \rightarrow Aut_K((E,A,\tau),(F,\sigma)) \rightarrow 1 \]
That the tori in $U_{E,A,\tau}$ correspond to elements in the kernel of the map from $H^1(N)$ to $H^1(U_{E,A,\tau})$ is equivalent to the claim that there is a unique element of $H^1(Z)$ whose image is in this kernel so that we have an isomorphism from this kernel in $H^1(N)$ to the above kernel in $Aut_K((E,A,\tau),(F,\sigma))$. This claim about the elements of $H^1(Z)$ follows by observing that elements of $H^1(Z)$ are describing the discriminant of the unitary group.
\end{remark}
